---
title:  |
  | **PHHA**
  | <span style='font-size: 18pt'>*Nucleotide diversity via angsd*</span>
output: 
  html_document:
    theme: cosmo
    highlight: tango
---

```{r, echo=FALSE}
library(knitr)
opts_chunk$set(echo = F, message = F, warning = F, fig.width = 8, fig.height = 8, fig.align = 'center')
```

First creating a downsized VCF. Since filtering on individuals for pop. gen dataset has already occurred, I'm going ahead and removing the same individuals here. I'm doing the same first step as with pop gen filtering **without** thinning (i.e. I'm keeping all SNPs on a given contig).

```{bash, echo=TRUE, eval=FALSE}
vcftools \
--gzvcf PHHA.vcf.gz \
--remove-indels \
--min-alleles 2 \
--max-alleles 2 \
--minQ 30 \
--remove indmiss85.txt \
--remove-filtered-all \
--recode \
--recode-INFO-all \
--out PHHA.bi.q30.i85
```

Produces file **PHHA.bi.q30.i85.recode.vcf**

Next, doing the same steps as pop. gen filtering to remove high-depth loci.

```{bash, echo=TRUE, eval=FALSE}
vcftools \
--vcf PHHA.bi.q30.i85.recode.vcf \
--site-mean-depth \
--out PHHA.bi.q30.i85
```

```{bash, echo=TRUE, eval=FALSE}
vcftools \
--vcf PHHA.bi.q30.i85.recode.vcf \
--freq2 \
--out PHHA.bi.q30.i85
```

Produces files called **PHHA.bi.q30.i85.ldepth.mean** and **PHHA.bi.q30.i85.frq**

Calculating missingness-adjusted depth and extracting loci to keep (using the same max threshold of **12**.

```{r, eval=FALSE, echo=TRUE}
var_depth <- read_delim("../diversity/PHHA.bi.q30.i85.ldepth.mean", delim = "\t",
           col_names = c("chr", "pos", "mean_depth", "var_depth"), skip = 1)

a_freq <- read_delim("../diversity/PHHA.bi.q30.i85.frq", delim = "\t",
                     col_names = c("chr", "pos", "nalleles", "nchr", "a1", "a2"), skip = 1) %>%
  mutate(fmiss = (max(nchr)-nchr)/max(nchr))

full_depth <- full_join(var_depth, a_freq, by = c("chr", "pos")) %>%
  mutate(depth_adj = mean_depth*(1/(1-fmiss)))

snp_set <- filter(full_depth, depth_adj < 12) %>%
  select(chr, pos)

write_delim(snp_set, '../diversity/pi_sites_maxdp12.tsv', delim = '\t', col_names = F)
```

Nucleotide diversity is also sensitive to missing data, and for simplicity sake, I'm going to use the same missingness threshold as in the pop. gen filtering (loci occurring in **>= 70%** of individuals). I can filter on both depth and missingness to produce a final VCF for nucleotide diversity. Note that this loci set is the same as the pop. g set minus **thinning** and **MAF** filters.

```{bash, echo=TRUE, eval=FALSE}
vcftools \
--vcf PHHA.bithin.q30.i85.recode.vcf \
--positions pi_sites_maxdp12.tsv \
--max-missing 0.7 \
--remove-filtered-all \
--recode \
--recode-INFO-all \
--out PHHA.bi.q30.i85.maxdp12.a70
```

Produces a final VCF called **PHHA.bi.q30.i85.maxdp12.a70.recode.vcf**

Next bit is setting everything up for `angsd`. The general idea is to create "population lists" (which are just text files listing individual .bam files for a given population) that will get passed through a custom `angsd` pipeline to calculate a single estimate of $\pi$ (nucleotide diversity). Because we have different sampling numbers for each population, the way that I test for robust estimates is to look at $\pi$ with rarefaction. In other words, we'll compare $\pi$ across populations based on calculations using 3, 6, 8, and 10 individuals per population. We also need to understand how sensitive a single downsampled calculation is (e.g. there will be variation in $\pi$ based on which 3 individuals get pulled from a population of 12), so we'll use bootstrapping. Therefore, far all 24 populations, I'll create population lists for each of these downsampling targets (4), and if the sampling for a population is large enough, I'll create up to 99 unique individual sets (or however many unique sets are possible given the total population sampling).

First step is to get a list of all individual IDs from which we will generate these sets. It's helpful to use the `.imiss` output from `vcftools` in case we want to go back and look at coverage (spefor individuals that might be disproportionately skewing $\pi$ estimates.

```{bash, echo=TRUE, eval=FALSE}
vcftools \
--vcf PHHA.bi.q30.i85.maxdp12.a70.recode.vcf \
--missing-indv \
--out PHHA.bi.q30.i85.maxdp12.a70
```

Produces a file called **PHHA.bi.q30.i85.maxdp12.a70.imiss**

Setting up a file to subsample from (`phha_ids`) and summary of pop. sample totals (`per_pop`).

```{r}
phha_ids <- read_delim("../diversity/PHHA.bi.q30.i85.maxdp12.a70.imiss", delim = "\t",
                       col_names = c("ind", "ndata", "nfiltered", "nmiss", "fmiss"), skip = 1) %>%
  mutate(Pop = str_split_i(ind, '_', 2))

per_pop <- group_by(phha_ids, Pop) %>%
  summarize(n = n(),
            cov = mean(fmiss))
```

Going to make some directories where these population lists will be organized

```{bash, echo=TRUE, eval=FALSE}
mkdir ../diversity/popLists_3per
mkdir ../diversity/popLists_6per
mkdir ../diversity/popLists_8per
mkdir ../diversity/popLists_10per
```

Now generating all population list (***NEED TO GO BACK AND FUNCTIONALIZE ALL THIS LATER FOR A SINGLE DOWNSAMPLE VALUE, BUT DOING MANUALLY/LONG WAYS FOR NOW***)

```{r}
pops_6per <- filter(per_pop, n > 5.5)$Pop
pops_8per <- filter(per_pop, n > 7.5)$Pop
pops_10per <- filter(per_pop, n > 9.5)$Pop

inds_6per <- filter(phha_ids, Pop %in% pops_6per)
inds_8per <- filter(phha_ids, Pop %in% pops_8per)
inds_10per <- filter(phha_ids, Pop %in% pops_10per)

# generating pop lists for 6 individuals per
for (i in 1:length(unique(inds_6per$Pop))) {
  pop = filter(inds_6per, Pop == unique(inds_6per$Pop)[i])
  total = seq(1, nrow(pop))
  combos = combn(total, 6, simplify = F)
  if(length(combos) < 100) {
    for (j in 1:length(combos)){
      chosen = slice(pop, combos[[j]])
      filename = paste0('../diversity/popLists_6per/',
                        unique(inds_6per$Pop)[i], '_', as.character(j), '.txt')
      writeLines(chosen$ind, filename)
    }} else{
      for (k in 1:99) {
        samp6 = sample(total, size = 6, replace = F)
        chosen = slice(pop, samp6)
        filename = paste0('../diversity/popLists_6per/',
                          unique(inds_6per$Pop)[i], '_', as.character(k), '.txt')
        writeLines(chosen$ind, filename)
      }}
}

# generating pop lists for 8 individuals per
for (i in 1:length(unique(inds_8per$Pop))) {
  pop = filter(inds_8per, Pop == unique(inds_8per$Pop)[i])
  total = seq(1, nrow(pop))
  combos = combn(total, 8, simplify = F)
  if(length(combos) < 100) {
    for (j in 1:length(combos)){
      chosen = slice(pop, combos[[j]])
      filename = paste0('../diversity/popLists_8per/',
                        unique(inds_8per$Pop)[i], '_', as.character(j), '.txt')
      writeLines(chosen$ind, filename)
    }} else{
      for (k in 1:99) {
        samp8 = sample(total, size = 8, replace = F)
        chosen = slice(pop, samp8)
        filename = paste0('../diversity/popLists_8per/',
                          unique(inds_8per$Pop)[i], '_', as.character(k), '.txt')
        writeLines(chosen$ind, filename)
      }}
}

# generating pop lists for 10 individuals per
for (i in 1:length(unique(inds_10per$Pop))) {
  pop = filter(inds_10per, Pop == unique(inds_10per$Pop)[i])
  total = seq(1, nrow(pop))
  combos = combn(total, 10, simplify = F)
  if(length(combos) < 100) {
    for (j in 1:length(combos)){
      chosen = slice(pop, combos[[j]])
      filename = paste0('../diversity/popLists_10per/',
                        unique(inds_10per$Pop)[i], '_', as.character(j), '.txt')
      writeLines(chosen$ind, filename)
    }} else{
      for (k in 1:99) {
        samp10 = sample(total, size = 10, replace = F)
        chosen = slice(pop, samp10)
        filename = paste0('../diversity/popLists_10per/',
                          unique(inds_10per$Pop)[i], '_', as.character(k), '.txt')
        writeLines(chosen$ind, filename)
      }}
}

# generating pop lists for 10 individuals per
for (i in 1:length(unique(phha_ids$Pop))) {
  pop = filter(phha_ids, Pop == unique(phha_ids$Pop)[i])
  total = seq(1, nrow(pop))
  combos = combn(total, 3, simplify = F)
  if(length(combos) < 100) {
         for (j in 1:length(combos)){
           chosen = slice(pop, combos[[j]])
           filename = paste0('../diversity/popLists_3per/',
                             unique(phha_ids$Pop)[i], '_', as.character(j), '.txt')
           writeLines(chosen$ind, filename)
         }} else{
         for (k in 1:99) {
           samp3 = sample(total, size = 3, replace = F)
           chosen = slice(pop, samp3)
           filename = paste0('../diversity/popLists_3per/',
                             unique(phha_ids$Pop)[i], '_', as.character(k), '.txt')
           writeLines(chosen$ind, filename)
         }}
}
```

Reminder to self that you can move whole directories recursivley via `scp` with `-r`.

Next need to generate a sites file via `vcftools`

```{bash, echo=TRUE, eval=FALSE}
vcftools \
--vcf PHHA.bi.q30.i85.maxdp12.a70.recode.vcf \
--kept-sites \
--out PHHA.bi.q30.i85.maxdp12.a70
```

Produces a file called **PHHA.bi.q30.i85.maxdp12.a70.kept.sites**

Then need to remove first line with column headers.

```{bash, echo=TRUE, eval=FALSE}
sed -i '1d' PHHA.bi.q30.i85.maxdp12.a70.kept.sites
```

Get into the angsd environment. Should have directions elsewhere to help with this, but for now doing the things I need.

```{bash, echo=TRUE, eval=FALSE}
startConda # personal alias for running conda on ponderosa
conda activate angsd
angsdPath # personal alias for fixing a gsl path
```

Next need to index the sites file.

```{bash, echo=TRUE, eval=FALSE}
~/angsd/angsd sites index PHHA.bi.q30.i85.maxdp12.a70.kept.sites
```

```{bash, echo=TRUE, eval=FALSE}
#!/usr/bin/env bash
# pipeline for a single population, used as workhorse in diversity_parallel_batch.sh
# Run SAF -> realSFS -> saf2theta -> thetaStat -> extract mean pi for one population
# Usage: ./pipeline_for_one_population.sh /path/to/poplist.txt /abs/path/to/outdir

set -u
# --------- USER-CONFIGURE as needed ----------
ANGSD="${ANGSD:-$HOME/angsd}"   # path to angsd installation (ex: $HOME/angsd)
SITES="${SITES:-../vcf/pi_filtering/PHHA.bi.q30.i85.maxdp12.a70.kept.sites}"
ANC="${ANC:-../assembly/PHHA_ref.fa}"
MAXITER="${MAXITER:-200}"
FOLD="${FOLD:-1}"   # 1 for folded spectra (ancestral state unknown), 0 for unfolded (ancestral known)
P_ARG="${P_ARG:-1}"   # threads passed to angsd commands, most efficient at 1 if using de novo reference
KEEP_INTERMEDIATES="${KEEP_INTERMEDIATES:-0}"   # 0 -> remove intermediates, 1 -> keep them
# ----------------------------------------------

poplist="$1"          # path to the bam list (e.g. pop_lists/B0.BM_fmiss60.rare12.txt)
outdir="$2"           # absolute or relative output directory (will be created if missing)
logdir="${outdir}/logs"
mkdir -p "$outdir" "$logdir"

# Resolve popname: basename without .txt
popname="$(basename "$poplist" .txt)"

timestamp() { date +"%Y-%m-%d %H:%M:%S"; }

logfile="$logdir/${popname}.pipeline.log"
echo "[$(timestamp)] START $popname" >> "$logfile"

# If final pi already exists, skip
pi_out="${outdir}/${popname}.pi.txt"
if [ -f "$pi_out" ]; then
    echo "[$(timestamp)] SKIP $popname (already has $pi_out)" >> "$logfile"
    exit 0
fi

set -o pipefail

# ------------------------
# Step 1: ANGSD doSaf
# ------------------------
saf_prefix="${outdir}/${popname}"
echo "[$(timestamp)] Running ANGSD -doSaf for $popname" >> "$logfile"
"$ANGSD/angsd" -bam "$poplist" \
    -doSaf 1 \
    -anc "$ANC" \
    -GL 1 \
    -P "$P_ARG" \
    -sites "$SITES" \
    -out "$saf_prefix" \
    >> "$logfile" 2>&1
saf_idx="${saf_prefix}.saf.idx"
if [ ! -f "$saf_idx" ]; then
    echo "[$(timestamp)] ERROR: $saf_idx not created - aborting $popname" >> "$logfile"
    exit 2
fi

# ------------------------
# Step 2: realSFS 1D SFS with full convergence checking
# ------------------------
echo "[$(timestamp)] Running realSFS for $popname" >> "$logfile"
cnvlog="$outdir/${popname}.sfs.cnvrglog"
"$ANGSD/misc/realSFS" "$saf_idx" \
    -maxIter "$MAXITER" \
    -P "$P_ARG" \
    -fold "$FOLD" \
    > "${outdir}/${popname}.sfs" \
    2> "$cnvlog"
if [ ! -s "${outdir}/${popname}.sfs" ]; then
    echo "[$(timestamp)] ERROR: SFS not created for $popname" >> "$logfile"
    exit 3
fi

# Check for early EM-break
if grep -q "Breaking EM" "$cnvlog"; then
    em_line=$(grep "Breaking EM" "$cnvlog" | tail -n 1)
    em_iter=$(echo "$em_line" | grep -o "iter:[0-9]*" | cut -d: -f2)
    em_prev=$(grep "lik\[" "$cnvlog" | tail -n 1)
    em_diff=$(echo "$em_prev" | grep -o "diff=[0-9.eE+-]*" | cut -d= -f2)


echo "[$(timestamp)] Convergence reached early at iter=$em_iter with diff=$em_diff" >> "$logfile"
converged=1


else
    # Extract final diff
    last_diff=$(grep -o "diff=[0-9.eE+-]*" "$cnvlog" | tail -n 1 | cut -d= -f2)
    if [ -z "$last_diff" ]; then
        echo "[$(timestamp)] ERROR: Could not parse diff" >> "$logfile"
        exit 4
    fi
    echo "[$(timestamp)] Final SFS diff at maxIter was $last_diff" >> "$logfile"


    # Numeric threshold check
    awk -v v="$last_diff" 'BEGIN { exit !(v < 1e-6) }'
    if [ $? -eq 0 ]; then
        echo "[$(timestamp)] SFS convergence successful at maxIter (diff=$last_diff)" >> "$logfile"
        converged=1
    else
        echo "[$(timestamp)] ERROR: SFS failed to converge at maxIter (final diff=$last_diff)" >> "$logfile"
        exit 5
    fi
fi

# ------------------------
# Step 3: saf2theta
# ------------------------
echo "[$(timestamp)] Running saf2theta for $popname" >> "$logfile"
"$ANGSD/misc/realSFS" saf2theta "$saf_idx" \
    -sfs "${outdir}/${popname}.sfs" \
    -fold "$FOLD" \
    -outname "${outdir}/${popname}" \
    >> "$logfile" 2>&1
theta_idx="${outdir}/${popname}.thetas.idx"
if [ ! -f "$theta_idx" ]; then
    echo "[$(timestamp)] ERROR: $theta_idx not created" >> "$logfile"
    exit 6
fi

# ------------------------
# Step 4: thetaStat do_stat -> write to .thetas.idx.pestPG
# ------------------------
pestpg="${outdir}/${popname}.thetas.idx.pestPG"
echo "[$(timestamp)] Running thetaStat do_stat for $popname" >> "$logfile"
"$ANGSD/misc/thetaStat" do_stat "$theta_idx" -win 1 -step 1 > "$pestpg" 2>> "$logfile"
if [ ! -s "$pestpg" ]; then
    echo "[$(timestamp)] ERROR: $pestpg missing or empty" >> "$logfile"
    exit 7
fi

# ------------------------
# Step 5: extract mean pi (column 5)
# ------------------------
echo "[$(timestamp)] Extracting mean pi for $popname" >> "$logfile"
mean_pi=$(awk 'NR>1 {sum += $5; n++} END {if(n>0) printf "%.8f", sum/n; else print "NA"}' "$pestpg")
echo "$popname,$mean_pi" >> "$pi_out"
echo "[$(timestamp)] mean_pi=$mean_pi for $popname" >> "$logfile"

# ------------------------
# Step 6: cleanup intermediates if requested
# ------------------------
if [ "$KEEP_INTERMEDIATES" -eq 0 ]; then
    echo "[$(timestamp)] Cleaning up intermediates for $popname" >> "$logfile"
    rm -f "${saf_prefix}.saf" "${saf_prefix}.saf.idx" \
          "${saf_prefix}.sfs" "${saf_prefix}.sfs.cnvrglog" \
          "${saf_prefix}.thetas.gz" "${saf_prefix}.thetas.idx" \
          "${saf_prefix}.saf.gz" "${saf_prefix}.saf.pos.gz" "${saf_prefix}.arg" \
          "$pestpg"
fi

echo "[$(timestamp)] DONE $popname (pi=$mean_pi)" >> "$logfile"
exit 0
```

One thing to keep in mind is that the log files this pipeline produces are relatively large (`angsd` will give you a ton of repetitive output, especially if using a de novo reference with hundreds of thousands of contigs). Easist thing for debugging is to look at lines with timestamps (they should let you know if a certain command ran or hit an error). For bracketed time stamps that this script produces, this regex works:

```{bash, echo=TRUE, eval=FALSE}
grep -E '\[[0-9]{4}-[0-9]{2}-[0-9]{2}.*[0-9]{2}:[0-9]{2}:[0-9]{2}\]' WH_9.pipeline.log | less
```

```{r, fig.height=4, fig.width=6}
col24 <- c("#6c2733", "#905f33", "#7b39d5", "#bce333", "#cd4cc3",
           "#49571A", "#74398b", "#D49202", "#0051FF", "#1cf1a3",
           "#7ccfd2", "#d04675", "#154975", "#6690bd", "#4ed31b",
           "#d1b3d8", "#77d898", "#fa756b", "#cccc9c", "#FCB61E",
           "#cb8985", "#FFE508", "#e72525", "#0A7D22")

pi_3per <- read.csv('../diversity/pi_summary_3per.csv') %>%
  rename(ID = pop) %>%
  mutate(Pop = str_split_i(ID, '_', 1),
         run = '3per')

pi_6per <- read.csv('../diversity/pi_summary_6per.csv') %>%
  rename(ID = pop) %>%
  mutate(Pop = str_split_i(ID, '_', 1),
         run = '6per')

div_3per <- ggplot(pi_6per, aes(x = Pop, y = pi,
                           group = Pop, fill = Pop, color = Pop)) +
  geom_jitter(pch = 21,
              size = 1,
              width = 0.2) +
  geom_boxplot(outliers = F,
               alpha = 0.1,
               linewidth = 0.25) +
  scale_color_manual(values = darken(col24[-16], 0.3)) +
  scale_fill_manual(values = col24[-16]) +
  theme_bw() +
  theme(legend.position = 'none')

div_3per
```

```{r, fig.width=15, fig.height=3}
library(forcats)

allRuns <- rbind(pi_3per, pi_6per) %>%
  mutate(pop_run = paste0(Pop, '.', run),
         Pop = fct_reorder(as.factor(Pop), pi, .fun = mean, .desc = T))

mpg_ordered <- mpg %>%
  mutate(cyl = fct_reorder(as.factor(cyl), hwy, .fun = mean, .desc = TRUE))


design = "
ABCDEFGHIJKLMNOPQRSTUVWX
"


div_all <- ggplot(allRuns, aes(x = pop_run, y = pi, fill = run, color = run)) +
  geom_jitter(pch = 21,
              size = 1,
              width = 0.25,
              alpha = 0.4) +
  # geom_boxplot(outliers = F,
  #              alpha = 0.1,
  #              linewidth = 0.25, width = 0.25) +
  stat_summary(
    fun = "mean", 
    geom = "errorbar", 
    aes(ymax = after_stat(y), ymin = after_stat(y)), 
    width = 0.85, # Controls the length of the line
    color = "red", 
    linetype = "solid",
    linewidth = 0.5
  ) +
  # facet_grid(project~Pop, scales = 'free_x', switch = 'x', space = 'free') +
  # facet_grid2(project~Pop, scales = 'free_x', switch = 'x', space = 'free',
  #             render_empty = F) +
  facet_manual(vars(Pop), design = design,
               scales = 'free_x', ) +
  scale_color_manual(values = darken(col24[7:8], 0.3)) +
  scale_fill_manual(values = col24[7:8]) +
  guides(color = guide_legend(override.aes = list(size = 5, alpha = 1,
                                                  stroke = 1),
                              title = "Ind. per calculation"),
         fill = guide_legend(title = "Ind. per calculation")) +
  labs(y = 'Nucleotide diversity (\u03c0)') +
  theme_minimal() +
  theme(legend.position = 'bottom',
        # axis.text.x = element_text(angle = 90),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 18,
                                    face = 'bold'),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        panel.spacing.x = unit(1, "pt"),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 18,
                                    face = 'bold'),
        axis.text.x = element_blank(),
        panel.border = element_rect(linewidth = 0.25, color = 'black',
                                    fill = 'transparent'),
        panel.background = element_rect(fill = 'white'))

div_all
```

